---
title: "Near_Earth_Comet_Preprocessing"
author: "Troy Allen"
date: "2024-06-05"
output: html_document
---

Let's load in the data and some necessary packages first.
```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(glmnet)

comet_data= read.csv("C:/Users/Troy/Downloads/Near-Earth_Comets_-_Orbital_Elements_20240605.csv")
comet_data
```

Checking for missing values:
```{r}
# Find nan values
missing_values <- comet_data %>% 
  summarise(across(everything(), ~ sum(is.na(.)))) %>% 
  gather(key = "Column", value = "MissingValues")

# Print nan values
missing_values
```

We can see the missing values are as follows:

A1 = 122 missing values
A2 = 119 missing values
A3 = 150 missing values
DT = 156 missing values

Obviously can't drop all of these rows due to the sheer number of missing values we have.

```{r}
# Print out the 4 problematic columns
print(comet_data %>% select(A1..AU.d.2., A2..AU.d.2., A3..AU.d.2., DT..d.))
```
Dropping the columns A3 and DT seems like the best option for now. Currently dropping A1/A2 as well since I wanted to see what Corey has to say about the importance of these columns in our models.

```{r}
# Drop the 'A3' and 'DT' columns
comet_data_no_A3DT <- comet_data %>%
  select(-c(A1..AU.d.2., A2..AU.d.2.,A3..AU.d.2., DT..d.))

# Print the dataset without A3/DT
comet_data_no_A3DT
```

Let's check for outliers now.

```{r}
# Filter out non-numeric columns
numeric_cols <- select_if(comet_data_no_A3DT, is.numeric)

# Calculate Z-scores for each numeric column
z_scores <- numeric_cols %>% 
  mutate(across(everything(), scale))

# Define threshold
threshold <- 2

# Identify outliers based on Z-score
outliers <- comet_data_no_A3DT %>% 
  filter(rowSums(abs(z_scores) > threshold, na.rm = TRUE) > 0)

print(outliers)
```

```{r}
# Filter the outlier columns based on threshold
outlier_columns <- apply(abs(z_scores) > threshold, 1, function(x) colnames(comet_data_no_A3DT)[x])

# Print which columns the outlier lies in and the entire row associated with outlier
for(i in 1:nrow(outliers)) {
  if(length(outlier_columns[[i]]) > 0) {
    cat("Row:", i, "\n")
    cat("Outlier columns:", paste(outlier_columns[i], collapse = ", "), "\n")
    print(outliers[i, ])
  }
}

```
Wanted to compare each to the mean and stddev of each column.

```{r}
# Filter out numeric columns
numeric_data <- comet_data_no_A3DT[, sapply(comet_data_no_A3DT, is.numeric)]

# Calculate mean of each column
column_means <- colMeans(numeric_data, na.rm = TRUE)

# Calculate standard deviation of each column
column_stddev <- sapply(numeric_data, sd, na.rm = TRUE)

# Print mean and standard deviation of each column
for (i in seq_along(column_means)) {
  cat("Column:", names(column_means)[i], "\n")
  cat("Mean:", column_means[i], "\n")
  cat("Standard Deviation:", column_stddev[i], "\n\n")
}
```

Creating a standardized version of the data in case it is needed.

```{r}
# Get numeric columns from cleaned data
numeric_comet_data <- comet_data_no_A3DT[, sapply(comet_data_no_A3DT, is.numeric)]

# Get mean + sd
means <- colMeans(numeric_comet_data)
sds <- apply(numeric_comet_data, 2, sd)

# Standardize each column and print
standardized_comet_data <- scale(numeric_comet_data, center = means, scale = sds)
standardized_comet_data

```

Some simple plots based on Eccentricity, perihelion distance, and orbital elements:

```{r}
ggplot(comet_data_no_A3DT, aes(x = e, y = i..deg.)) +
  geom_point() +
  labs(x = "Eccentricity", y = "Inclination (deg)", title = "Scatter Plot of Eccentricity vs. Inclination")
```
```{r}
ggplot(comet_data_no_A3DT, aes(x = e, y = q..AU.)) +
  geom_point() +
  labs(x = "Eccentricity", y = "Perihelion Distance (q AU)", title = "Scatter Plot of Eccentricity vs. Perihelion Distance") 
```

```{r}
# Filter orbital elements 
orbital_elements <- comet_data[, c("q..AU.", "Epoch..TDB.", "TP..TDB.", "e", "i..deg.", "w..deg.", "Node..deg.")]

# Create a scatterplot matrix showing relationship between q and other orbital elements
pairs(orbital_elements, pch = 21, bg = "skyblue")
```

```{r}
write.csv(comet_data_no_A3DT, "comet_data_cleaned.csv", row.names = FALSE)
```

